Graphical Tests for Power Comparison of Competing Designs

### 2012-03-25 experiment.tex
changed cls file to refer to pdf not eps, and spelling of euclidian in experiment.tex

\section{Experimental Setup}
Describe the experimental setup: polar charts versus euclidean - with and without reference lines.
Figure showing all four types using the same dataset. 
Discuss construction of each one. Discuss construction of null chart.

The goal of this experiment is to understand the efficiency of displaying information using euclidian and polar coordinates. Efficiency is the deviation from independence, which can be simulated by taking permutations of the original dataset. Data concerning wind patterns and time between flights at SEA airport is used for these charts. The lineup chart is composed of 20 graphics with all of the exact same characteristics except for the data. One of the 20 graphics will be using original airport data while the other 19 will be permutations of that data. Efficiency of the chart types can be inferred by how accurately individuals pick out the plot with the original data. There are other variables which are introduced to further our understanding of the way we perceive euclidian and polar coordinates. All combinations of each level of the variables are involved in the experiment for both polar and euclidian coordinates. The graphics in each chart will all have exactly the same level of the variables with the only difference being the data.

The other variables which are monitored are existence of reference line, placement of x-axis, and sample size. It is possible that one of the chart types benefits more from additional components such as reference lines. The polar and euclidian null charts will be shown either with or without a white reference line. The line is placed at slightly over 50 percent of the y axis. Additionally, the x-axis has four placements. The original dataset contains a characteristic wave pattern, which for euclidian coordinates, may be a large factor contributing to the perceived effectiveness of the graphic. Characteristic patterns such as the wave pattern may be perceived in a different way when viewed in polar coordinates. When the x-axis is is shifted the look of the pattern changes. The wave pattern is broken up for euclidian coordinates but for polar coordinates is only shifted around the axis. Finally, the amount of information necessary for detection as well as correctness is a measure for efficiency of the charts so the sample size which was taken from the original dataset is of interest. The original data set contained XXX,XXX observations from which 2, 4, 6, 8, 10, and 24 percent of the observations were sampled. Small sample sizes are easier to obtain because of time and money constraints. It would be nice to know if there is one type of chart which can show data trends using less information.

 All combinations of these variables were repeated for plots with a reference line and without a reference line, as shown in figure \ref{layouts}.This figure shows examples of graphics showing the actual data, which in the experiment would be randomly positioned among 19 null plots. In total there are XXX unique charts which represent all different combinations of sample size, coordinate type, x-axis placement, and whether or not there is a reference line. The charts are all assigned a difficulty level, which is based on sample size, on a scale from zero to four. There are two plots assigned a difficulty level of zero. These two charts, one polar and one Euclidian, were made from computer generated data and intended to be at the level where an individual, if he or she is not purely guessing, should be able to find the correct plot.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=0.4\linewidth]{Euclidian_NoLine.pdf} \hspace{0.05\linewidth}
   \includegraphics[width=0.4\linewidth]{Euclidian_Line.pdf} \\
   \includegraphics[width=0.475\linewidth]{Polar_NoLine.pdf} 
   \includegraphics[width=0.475\linewidth]{Polar_Line.pdf}  
   \caption{All four layouts of charts The first row shows the data in polar coordinates and the second row shows the data in euclidian coordinates. The first column in both rows show the plots without any reference lines. The second column in both rows have a white reference line added. }
   \label{layouts}
\end{figure}

@@

Most generally this experiment can be paralleled to the pie chart versus barchart discussion. The Euclidian coordinate plot is made up of bars *whose?* areas are colored proportionally according to the data. The polar coordinate plot is made up of pie slices which are also colored proportionally according to the data. Viewers of these charts will visually decode the areas of the bars and pie slices and make a judgement based on their decoding. How accurate this judgment is is left to be discussed. According to (Kosslyn, Stephen, Graph Design for the Eye and Mind, Oxford University Press, 2006, p. 40), area is usually perceived as a function of what the actual area is. This function is the area raised to an exponent of approximately 0.8 and then multiplied by a constant. However, when bars are parallel, the relative height of these bars is perceived very accurately. From this information, one could infer that the Euclidian coordinate plot may gain efficiency based on the height of the colored bars.

@@

>>>>

\section{Experimental Setup}
Describe the experimental setup: polar charts versus Euclidean - with and without reference lines.
Figure showing all four types using the same dataset. 
Discuss construction of each one. Discuss construction of null chart.

The goal of this experiment is to understand the efficiency of displaying information using Euclidean and polar coordinates. Efficiency is the deviation from independence, which can be simulated by taking permutations of the original dataset. Data concerning wind patterns and time between flights at SEA airport is used for these charts. The lineup chart is composed of 20 graphics with all of the exact same characteristics except for the data. One of the 20 graphics will be using original airport data while the other 19 will be permutations of that data. Efficiency of the chart types can be inferred by how accurately individuals pick out the plot with the original data. There are other variables which are introduced to further our understanding of the way we perceive Euclidean and polar coordinates. All combinations of each level of the variables are involved in the experiment for both polar and Euclidean coordinates. The graphics in each chart will all have exactly the same level of the variables with the only difference being the data.

The other variables which are monitored are existence of reference line, placement of x-axis, and sample size. It is possible that one of the chart types benefits more from additional components such as reference lines. The polar and Euclidean null charts will be shown either with or without a white reference line. The line is placed at slightly over 50 percent of the y axis. Additionally, the x-axis has four placements. The original dataset contains a characteristic wave pattern, which for Euclidean coordinates, may be a large factor contributing to the perceived effectiveness of the graphic. Characteristic patterns such as the wave pattern may be perceived in a different way when viewed in polar coordinates. When the x-axis is is shifted the look of the pattern changes. The wave pattern is broken up for Euclidean coordinates but for polar coordinates is only shifted around the axis. Finally, the amount of information necessary for detection as well as correctness is a measure for efficiency of the charts so the sample size which was taken from the original dataset is of interest. The original data set contained XXX,XXX observations from which 2, 4, 6, 8, 10, and 24 percent of the observations were sampled. Small sample sizes are easier to obtain because of time and money constraints. It would be nice to know if there is one type of chart which can show data trends using less information.

 All combinations of these variables were repeated for plots with a reference line and without a reference line, as shown in figure \ref{layouts}.This figure shows examples of graphics showing the actual data, which in the experiment would be randomly positioned among 19 null plots. In total there are XXX unique charts which represent all different combinations of sample size, coordinate type, x-axis placement, and whether or not there is a reference line. The charts are all assigned a difficulty level, which is based on sample size, on a scale from zero to four. There are two plots assigned a difficulty level of zero. These two charts, one polar and one Euclidean, were made from computer generated data and intended to be at the level where an individual, if he or she is not purely guessing, should be able to find the correct plot.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=0.4\linewidth]{Euclidian_NoLine.pdf} \hspace{0.05\linewidth}
   \includegraphics[width=0.4\linewidth]{Euclidian_Line.pdf} \\
   \includegraphics[width=0.475\linewidth]{Polar_NoLine.pdf} 
   \includegraphics[width=0.475\linewidth]{Polar_Line.pdf}  
   \caption{All four layouts of charts The first row shows the data in polar coordinates and the second row shows the data in euclidean coordinates. The first column in both rows show the plots without any reference lines. The second column in both rows have a white reference line added. }
   \label{layouts}
\end{figure}

@@

Most generally this experiment can be paralleled to the pie chart versus barchart discussion. The Euclidean coordinate plot is made up of bars *whose?* areas are colored proportionally according to the data. The polar coordinate plot is made up of pie slices which are also colored proportionally according to the data. Viewers of these charts will visually decode the areas of the bars and pie slices and make a judgement based on their decoding. How accurate this judgment is is left to be discussed. According to (Kosslyn, Stephen, Graph Design for the Eye and Mind, Oxford University Press, 2006, p. 40), area is usually perceived as a function of what the actual area is. This function is the area raised to an exponent of approximately 0.8 and then multiplied by a constant. However, when bars are parallel, the relative height of these bars is perceived very accurately. From this information, one could infer that the Euclidean coordinate plot may gain efficiency based on the height of the colored bars.

@@

Euclidean coordinates performed better when speaking of efficiency in terms of accuracy and speed. 76.92\% of the Euclidean charts shown resulted in the a correct identification of the real data while only 20.31\% of the polar charts did the same. A t-test comparing means for charts with polar coordinates versus charts with Euclidean coordinates resulted in a t statistic of 15.2013 with a p-value of 2.2e-16. Similarly, a t-test comparing means of time spent on charts with Euclidean coordinates versus charts with polar coordinates resulted in a p-value of 1.014e-11. The average of the log of time spent on each individual chart for charts with Euclidean coordinates is 3.53 minutes while it was 4.07 minutes for charts with polar coordinates. Euclidean coordinates resulted in a significantly more accurate identification of the real data set in significantly shorter time.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=3in]{turk4_time_perc_preds.pdf}  
   \caption{Values predicted from XXXX how to say itXXXX of accuracy for log of time taken and test parameter.}
   \label{accuracy_preds}
\end{figure}

Figure \ref{accuracy_preds} shows predicted accuracy levels. These predicted values were obtained by using R package lme4 and a mixed effect model explaining correct responses by type of chart and the log of time spent while accounting for different levels of accuracy for each individual who participated in the experiment. The predicted values show that there is an overall higher accuracy level for Euclidean coordinates when compared to polar coordinates. For both chart types, as the time spent increases, the predicted accuracy decreases.

### 2012-03-25 intro.tex
changed cls file to refer to pdf not eps, and spelling of euclidian in experiment.tex

Usually, we need simulation studies to be able to completely control for the signal strength in the data. 
For charts, this is not a very realistic scenario. The need for displaying information particularly efficiently or accurately usually arises from  a particular problem while analyzing a data set, from which we want to learn new insight. Often, it is difficult to simulate data in such a way that it realistically represents the particular problem, which in turn makes it dubious whether the results from a simulation study are actually applicable to the problem at hand, reminding us painfully of George Box's famous quote on models: ``essentially, all models are wrong, but some are useful." \citep[page=424]{box}

>>>>

Usually, we need simulation studies to be able to completely control for the signal strength in the data. 
For charts, this is not a very realistic scenario. The need for displaying information particularly efficiently or accurately usually arises from  a particular problem while analyzing a data set, from which we want to learn new insight. Often, it is difficult to simulate data in such a way that it realistically represents the particular problem, which in turn makes it dubious whether the results from a simulation study are actually applicable to the problem at hand, reminding us painfully of George Box's famous quote on models: ``essentially, all models are wrong, but some are useful.'' \citep[page=424]{box}
